Random Generation of Rich \\ Abstract Data Type Values

### 2019-01-31 paper/randomtesting.tex
Pass over section 2

In this section, we briefly introduce the common approach for automatically
deriving random data generators for ADTs in QuickCheck by using the information
found at the type-level.
%
%type-driven approach, along
%with its main drawbacks.
%
% Haskell is a strongly typed programming language with a powerful type system.
% %
% It lets programmers encode a considerable amount of information about the
% structure of their systems using data types that can be checked at compilation
% time.
% %
% One of its key aspects is the support for Algebraic Data Types (ADTs).
% %
% Essentially, an ADTs is a composite type defined by combining other types in
% terms of \textbf{sums} (also known as \emph{variant types}) and
% \textbf{products} (or tuples) of other data types.
% %
%
To exemplify this, and for illustrative purposes, let us consider the following
ADT definition to encode simple |Html| pages:
%
\begin{code}
data Html  =   Text    String
           pp  Single  String
           pp  Tag     String  Html
           pp  Join    Html    Html

\end{code}
%
The type |Html| allows to build pages via four possible constructions:
|Text|---which represents plain text values---, |Single| and |Tag|---which
represent singular and paired Html tags, respectively---, and |Join|---which
concats two Htmls pages one after another.
%
% In the previous definition, we declare |Html| as the sum of four possible
% constructions: |Text| represents plain text values. |Sing| and |Tag| represent
% singular and paired Html tags, and |Join| concats two expression one after
% another.
% %
% We only encode a very small subset of the actual Html specification for
% illustrative reasons.
% %
In Haskell, |Text|, |Single|, |Tag|, and |Join| are known as \emph{data
  constructors} (or constructors for short) and are used to distiguish which
variant of the ADT we are constructing.
%
Each data constructor is defined as a product of zero or more types known as
\emph{fields}.
%
For instance, |Text| has a field of type |String|, whereas |Join| two recusive
fields of type |Html|.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this representation, the example page \texttt{<html>hello<hr>bye</html>}
can be encoded as:
%
\begin{code}
Tag "html" (Join (Join
  (Text "hello") (Single "hr")) (Text "bye"))
\end{code}
%
% Additionally, we can define a function |render| to serialize |Html| values as
% follows:
% %
% \begin{code}
% render :: Html -> String
% render (Text t) = t
% render (Sing t) = "<" ++ t ++ ">"
% render (Join x y) = render x ++ render y
% render (Tag t x)
%   = "<" ++ t ++ ">" ++ render x ++ "</" ++ t ++ ">"
% \end{code}

@@

In order to generate random value of types involving ADTs, most approaches
require users to provide a random data generator for each one of them.
%
This is a cumbersome and error prone task that usually \emph{follows closely the
  ADT structure}.
%
For instance, consider the following definition of a \quickcheck random
generator for the type |Html|:
%
\begin{code}
genHtml = sized (\size ->
  if size == 0
  then frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString) ]
  else frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString)
  ,  (4,  Tag     <$> genString <*> smaller genHtml)
  ,  (3,  Join    <$> smaller genHtml <*> smaller genHtml) ])
\end{code} %$
%
We use the Haskell syntax |[]| and |(,)| for denoting lists and pairs of
elements, respecitvely, e.g., |[(1,2),(3,4)]| is a list of pairs of numbers.
%
The random generator |genHtml| is defined using \quickcheck's function |sized|
to paremetrize the generation process up to an external natural number known as
the \emph{generation size}---captured in the code with variable |size|.
%
This parameter is chosen by the user, and it is used to limit the maximum amount
of recursive calls that this random generator can perform and thus ensuring
the termination of the generation process.
%
When called with a positive generation size, this generator can pick to generate
among any |Html| data constructor of with an explicitly given \emph{generation
  frequency} that can be chosen by the user---in this example, 2, 1, 4 and 3 for
|Text|, |Single|, |Tag|, and |Join|, respectively.
%
When it picks to generate a |Text| or a |Single| data constructor, it also
generates a random |String| value using the standard QuickCheck generator
|genString|.% (\quickcheck provides standard random generators for most base types
% like |String|, |Int|, |Bool|, etc.)
\footnote{%
  The operators |<$>| and |<*>| are used in Haskell to combine values obtained
  from calling random generators and they are not particularly relevant for the
  point being made in this work.}
%$
On the other hand, when it picks to generate a |Join| constructor, it also
generates two independent random sub-expression recursively, decreasing the
generation size by a unit on each recursive invocation (|smaller genHtml|).
%
The case of random generation of |Tag| constructors follows analogously.
%
This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick only among terminal
data constructors, being |Text| and |Single| the only possible choices in our
particular case.

>>>>

In this section, we briefly introduce the common approach for automatically
deriving random data generators for ADTs in QuickCheck.
%by using the information
%found at the type-level.
%
%type-driven approach, along
%with its main drawbacks.
%
% Haskell is a strongly typed programming language with a powerful type system.
% %
% It lets programmers encode a considerable amount of information about the
% structure of their systems using data types that can be checked at compilation
% time.
% %
% One of its key aspects is the support for Algebraic Data Types (ADTs).
% %
% Essentially, an ADTs is a composite type defined by combining other types in
% terms of \textbf{sums} (also known as \emph{variant types}) and
% \textbf{products} (or tuples) of other data types.
% %
%
To exemplify this, and for illustrative purposes, let us consider the following
ADT definition to encode simple |Html| pages:
%
\begin{code}
data Html  =   Text    String
           pp  Single  String
           pp  Tag     String  Html
           pp  Join    Html    Html

\end{code}
%
The type |Html| allows to build pages via four possible constructions:
|Text|---which represents plain text values---, |Single| and |Tag|---which
represent singular and paired Html tags, respectively---, and |Join|---which
concats two Htmls pages one after another.
%
% In the previous definition, we declare |Html| as the sum of four possible
% constructions: |Text| represents plain text values. |Sing| and |Tag| represent
% singular and paired Html tags, and |Join| concats two expression one after
% another.
% %
% We only encode a very small subset of the actual Html specification for
% illustrative reasons.
% %
In Haskell, |Text|, |Single|, |Tag|, and |Join| are known as \emph{data
  constructors} (or constructors for short) and are used to distiguish which
variant of the ADT we are constructing.
%
Each data constructor is defined as a product of zero or more types known as
\emph{fields}.
%
For instance, |Text| has a field of type |String|, whereas |Join| has two
recursive fields of type |Html|.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this representation, the example page \texttt{<html>hello<hr>bye</html>}
can be encoded as:
%
\begin{code}
Tag "html" (Join (Join
  (Text "hello") (Single "hr")) (Text "bye"))
\end{code}
%
% Additionally, we can define a function |render| to serialize |Html| values as
% follows:
% %
% \begin{code}
% render :: Html -> String
% render (Text t) = t
% render (Sing t) = "<" ++ t ++ ">"
% render (Join x y) = render x ++ render y
% render (Tag t x)
%   = "<" ++ t ++ ">" ++ render x ++ "</" ++ t ++ ">"
% \end{code}

@@

In order to generate random ADTs values, most approaches require users to provide
a random data generator for each ADT definition.
%
This is a cumbersome and error prone task that usually \emph{follows closely
  ADTs structures}.
%
For instance, consider the following definition of a \quickcheck random
generator for the type |Html|:
%
\begin{code}
genHtml = sized (\size ->
  if size == 0
  then frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString) ]
  else frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString)
  ,  (4,  Tag     <$> genString <*> smaller genHtml)
  ,  (3,  Join    <$> smaller genHtml <*> smaller genHtml) ])
\end{code} %$
%
We use the Haskell syntax |[]| and |(,)| for denoting lists and pairs of
elements, respecitvely (e.g., |[(1,2),(3,4)]| is a list of pairs of numbers.).
%
The random generator |genHtml| is defined using \quickcheck's function |sized|
to paremetrize the generation process up to an external natural number known as
the \emph{generation size}---captured in the code with variable |size|.
%
This parameter is chosen by the user, and it is used to limit the maximum amount
of recursive calls that this random generator can perform and thus ensuring
the termination of the generation process.
%
When called with a positive generation size, this generator can pick to generate
among any |Html| data constructor with an explicitly \emph{generation frequency}
that can be chosen by the user---in this example, 2, 1, 4 and 3 for |Text|,
|Single|, |Tag|, and |Join|, respectively.
%
When it picks to generate a |Text| or a |Single| data constructor, it also
generates a random |String| value using the standard QuickCheck generator
|genString|.% (\quickcheck provides standard random generators for most base types
% like |String|, |Int|, |Bool|, etc.)
\footnote{%
  The operators |<$>| and |<*>| are used in Haskell to combine values obtained
  from calling random generators and they are not particularly relevant for the
  point being made in this work.}
%$
On the other hand, when it picks to generate a |Join| constructor, it also
generates two independent random sub-expression recursively, decreasing the
generation size by a unit on each recursive invocation (|smaller genHtml|).
%
The case of random generation of |Tag| constructors follows analogously.
%
This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick among terminal data
constructors, being |Text| and |Single| the only possible choices in our
particular case.
